---
title: "A Binary Logistic Regression Model To Identify The Key Factors Of Smoking Using 2016 GSS Data"
author: "AnanJiAng, Fang Jia"
date: "10/18/2020"
output: html_document
Code and data supporting this analysis is available at: ""
---
##Abstract

Smoking is a major health problem worldwide; it will increase the risk of having the disease. We used the 2016 General social survey on Canadians at Work and Home data to build a logistic regression model to predict whether an individual will smoke or not given age groups, gender, education level, stress levels, and a number of children in one household. This model might help in developing successful prevention programs for smokers.

##Introduction

We obtained the 2016 GSS data from 19609 respondents, along with five features and one response variable. We divided 19609 respondents into ten provinces within Canada and built a logistic regression model to predict whether individuals will smoke or not. By investigating the logistic regression model, we could tell which features impact an individual will smoke or not. In the following aspects, we will explain how we process the 2016 GSS data and how we interpret the result from the logistic regression model.o 1557.79. It already has increasing trend of stock price. 

##Data

The dataset we obtained is the 2016 General Social Survey(GSS). The target population is Canadians who have over 15 years of age at work and home but excluding residents of Yukon, Northwest Territories, and Nunavut as well as full-time residents of institutions. The total number of variables in the dataset is 2194. There are 19609 Canadians took this survey and recorded in the 2016 GSS dataset.We created a data frame by selecting six variables from the GSS dataset. The data frame contained five predictor variables (age group, gender, education level, stress levels, and a number of children in one household) and one responsible variable (people smoke or not).

The GSS program used two approaches to collect 2016 data, which were electronic questionnaire(EQ) and via computer-assisted telephone interviews(CATI). To reach desired respondents, paper introductory letters with a secure access code were sent to households. Once a household member confirmed his or her age was 15 years or over and provided the primary phone number belonged to the household, he or she would list the names, sex, and ages of all people in one household. (If the phone number did not belong to him or her, the case would be transferred to CATI.) Then someone from the particular household would be randomly selected as a respondent and kept on the survey. A respondent who is not on a roster will leave an email address or phone number to continue doing the survey. 

The overall response rate for 2016 GSS was 50.8%. This reflected that almost half of the respondents didn't provide desired responses. To increase the response rate, the GSS program decided to adjust non-response. Three types grouped Non-respondents' phone numbers:
Type 1: Non-respondents with some auxiliary information. (i.e., complete information of household member)
Type 2: Non-respondents with auxiliary information from various sources available to Statistics Canada.
Type 3: Non-respondents with no auxiliary information.
The adjustment was divided into three stages. The first stage focuses on non-respondents in type 3 within each stratum independently. The second stage mainly focuses on non-respondents in type 2. They had auxiliary information which was used to a model propensity to respond. The third stage adjusted non-respondents in type 1. After adjusting non-response, the non-response phone numbers were finally dropped.

The survey used stratified sampling method to stratify data into 10 provinces in Canada. Stratified sampling method divides total population into smaller groups or strata to complete the sampling process. The strata is formed based on some common characteristics in the population data. After dividing the population into strata, we can randomly select the sample proportionally. Usually, simple random sampling method is a good choice because it provides equal chance for respondents to take survey. However, let's consider if one province has only 300 residents while the other province has 2000 residents. We would like to use 2000 as sample size to make statistical prediction. There will occur under sample error in the province with 300 residents because we use 300 respondents do predict a 2000 sample. The outcome of analysis must be incorrect. Same as over sample error. Therefore, GSS data chose stratified sampling method was a better decision. Sample would be stratified by different groups so each group would have a corresponding sample size. This method is good for reducing over-sample or under-sample error. 19609 Canadians took this survey. In general, the dataset is useful to predict if Canadians will smoke or not but due to the restriction of only Canada data, we can't tell the whole world if the one will smoke or not who lived outside Canada. The dataset was also too old which was recorded on August 2 to December 23, 2016. If we can have some recent and future data to analysis and compare, we can predict a more precise outcome.

##Model

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(survey)
# considered smoking occasionally and daily as YES, count not at all smoking as NO, count don't know, refusal and not stated as NA
data<-gss
str(data)
ss <- data %>% 
  rowwise() %>% 
  mutate(smoking_or_not = case_when(
    smoking=="Occasionally" ~ "YES",
    smoking=="Daily"~ "YES",
    smoking=="Not at all" ~ "NO",
    smoking=="Don't know" ~ "NA",
    smoking=="Refusal" ~ "NA",
    smoking=="Not stated"~"NA",

  )) 
# remove the don't know, refusal and not stated of smoking, then name the data as ss2
  ss2 <-ss[!(ss$smoking_or_not=="NA"),]
ss2
# Making smoking equals to 1, not smoking equals to 0
ss2$smoking.type<-as.factor(ifelse(ss2$smoking_or_not=="YES","1","0")) 
# from the guidebook of dataset in the website, put each province's population into survey design. 
ss2$fpc<-ifelse(ss2$province=="Alberta",1737,ifelse(ss2$province=="Ontario",5069,
                                              ifelse(ss2$province=="Newfoundland and Labroador",1295,ifelse(ss2$province=="Prince Edward Island",630,ifelse(ss2$province=="Nova Scotia",1264,ifelse(ss2$province=="New Brunswick",1287,ifelse(ss2$province=="Quebec",3132,ifelse(ss2$province=="Manitoba",1274,ifelse(ss2$province=="Saskatchewan",1374,ifelse(ss2$province=="British Columbia",2547,0))))))))))
example.design.strs<-svydesign(id=~1,strata=~province, data=ss2, fpc=~fpc)
# Use above survey design to make a svyglm, with y is smoking and non-smoking, x is five variables.
svyglm.strs.logit <- svyglm(smoking.type ~ sex+number_of_children+level_of_stress+education_level+agegroup, example.design.strs, family="binomial")
summary(svyglm.strs.logit)
# All the coefficients of each x 
svyglm.strs.logit$coefficients
```
##Model

We built a logistic regression model to predict if people would smoke or not. A logistic regression model is a statistical model that, in its basic form, uses a logistic function to model a binary response variable. Four assumptions are using the logistic regression model. Firstly, the respondent variable must be binary. In this model, our binary response variable is people smoke or not smoke. Secondly, the observations of logistic regression must be independent. Each observation represents a person in one household. Thirdly, the independent variables must not be highly correlated with each other. We choose five independent variables, which are sex, number of children, stress levels, education levels, and age groups. These variables have no direct relationship with each other. Thirdly, there is a linearly related relationship between independent variables and the log odds. The logistic regression model is log(p/1-p) = β0+ β1(sexMale) + β2(number_of_childrenNone) + ... + β6(number_of_childrenTwo) + β7(level_of_stress don't know) + ... + β13(level_of_stressRefusal) + β14(education_levelNot stated) + ... + β18(education_levelValid skip) + β19(agegroup25 to 34 years) + ...+ β23(agegroup75 years and over). It shows there is a linear relationship between x and log odds.


The software we use is the function called svyglm. Svyglm is used to fit a generalized linear model to data from a complex survey design, with inverse-probability weighting and design-based standard errors.The logistic function is log(p/(1-p)) = β0 + β1X and we can use it to write our own logistic function: 
log(p/1-p) = β0+ β1(sexMale) + β2(number_of_childrenNone) + ... + β6(number_of_childrenTwo) + β7(level_of_stressDon't know) + ... + β13(level_of_stressRefusal) + β14(education_levelNot stated) + ... + β18(education_levelValid skip) + β19(agegroup25 to 34 years) + ...+ β23(agegroup75 years and over)

β is the coefficient that represents the change in log odds for every one unit increase in an X variable. The positive coefficients in our model represent more likely to smoke. The negative coefficients in our model represent more likely not to smoke. For example, the coefficient of age group 65 to 75 years is -0.592887419. So we can predict that people who have 65 to 74 years are less likely to smoke. The coefficient of extremely stressful is 0.679148743. Therefore, we predict that people who are extremely stressed right now are more likely to smoke. 

β0 = -3.3883:  When sex=Female, number_of_children = Four or more, level_of_stress = A bit stressful, education_level = Elementary, junior high school or high school, agegroup = 15 to 24 years, log odds of people smoke is equal to -3.3883.

β1 = 0.2475:  When individuals are male, and everything remains the same, log odds of people's smoke will increase by 0.2475.

β2-β6: It is the dummy variable of number_of_children. For example, a dummy variable is when one household has one child, the x equals 1, and the other x(one household has more than one child, zero children) equals 0.
β2 = 0.5776: When one household has zero number_of_children, the log odds of people's smoke will increase by 0.5776.
If a coefficient is negative, then log odds of people's smoke will decrease. The rest of all predictor variables and dummy variables work in the same way.

P-value is important to test if an independent variable has a significant effect on the dependent response variable. Usually, the testing variable is significant if p-value<0.05; otherwise, it is not significant. The smaller the p-value, the more significant the variable will decide if people smoke or not. For example, the p-value of the age group 35-44 years is 5.88e^-9. So we can predict if people will smoke or not according to check if their ages are 35-44 years. The p-value of the age group 45 to 54 years is 0.845, which means we can not predict this group of people will smoke or not. We observed that only two variables were not significant. They were level_of_stress don't know and agegroup45 - 54 years. In a word, if you meet a person who tells you he/she is 47 years old, you can not directly predict if he/she smokes. Instead, you need to ask more about his/her personal information, such as if he/she has two children so that you can know he/she is likely to smoke. The rest 21 dummy variables have a heavy influence on predicting if people smoke or not.

We can now make predictions from the estimates according to some situations. We can compute the effects for all of the predictor variables for a particular scenario, adding them up, and applying a logistic transformation. 
Consider for a scenario:
"Mike is a 25 years old university student who studies STA304. He's not at all stressed in this course because of receiving a nice score on his problem set."
To predict if Mike smokes or not, we first put him into the fitting variables: sex=male, level_of_stress=Not at all, education_level = university, age group = 25 to 34 years.
Then we use logistic regression function:
log(p/1-p) = -3.3883 + (sexMale=1)*0.2475 + (level_of_stressNot at all = 1)* 0.0949  + (education_levelUniversity=1)*0.5250 = -2.5209
In conclusion, the log odds of "if Mike smoke" = -2.5209. We can transfer it into probability = exp(-2.5209)=8.04%
So we can also conclude that Mike has a probability of 8.04% to smoke.
##Results
```{r,include=FALSE,echo=TRUE}
# create a bar plot with sex and (smoking, non-smoking)
ggplot(ss2, aes(x = sex, fill = smoking_or_not)) +
    geom_bar(position = "fill") +
    theme_classic() + 
  labs(x="gender",
       title= "Figure 1: The distribution of individuals smoking given female and male  ",
       subtitle = "Changes in individuals smoking given different gender ",
       cex.names = 0.9)+
  theme_bw()+theme(axis.text.x = element_text(angle = 30, hjust = 1, size=10,color="darkred"))
```

By analyzing figure 1, the number of males who choose to smoke more than the number of females chooses to smoke. In the model result, the sex of female counts as a baseline. The coefficient of the male in the logistic regression model is positive(0.247). When an individual's gender is male, and other variables remain the same, the odds of individual smoke is increased by 0.247. 

```{r,include=FALSE,echo=TRUE}
# create a bar plot with different level of stress and (smoking, non-smoking)
ggplot(ss2, aes(x = level_of_stress, fill = smoking_or_not)) +
    geom_bar(position = "fill") +
    theme_classic() + 
  labs(x="stress level",
       title= "Figure 2: The distribution of individuals smoking given eight stress levels  ",
       subtitle = "Changes in individuals smoking given different stress levels ",
       cex.names = 0.9)+
  theme_bw()+theme(axis.text.x = element_text(angle = 30, hjust = 1, size=10,color="darkred"))
```

From figure 2, the number of extreme stress people who smoke is the most among all eight stress levels.
The baseline for stress levels is a bit stressful. Don't know in stress level is not significant in evaluating people will smoke because the p-value is 0,97, which is larger than 0.05. We will not consider the don't know stress level when building the logistic regression model. The coefficient of not very stressful and refusal is negative(-0.1367 and -10.7448). When an individual is above two stress levels, the odds of people smoke will decrease. The rest of the five stress levels have a positive coefficient; it represents an increase in people's odds of smoking. 

```{r,include=FALSE,echo=TRUE}
# create a bar plot with different age groups and (smoking, non-smoking)
ggplot(ss2, aes(x = agegroup, fill = smoking_or_not)) +
    geom_bar(position = "fill") +
    theme_classic() + 
  labs(x="stress level",
       title= "Figure 3: The distribution of individuals smoking given seven agegroups",
       subtitle = "Changes in individuals smoking given different age range ",
       cex.names = 0.9)+
  theme_bw()+theme(axis.text.x = element_text(angle = 30, hjust = 1, size=10,color="darkred"))
```

From figure 3, the number of people in 25-34 years old and 45-64 years old choose to smoke approximately are the most among all seven age groups. The baseline for age groups is 15-24. However, the 45-54 age group is not significant to predict people will smoke because the p-value, which is 0.845, is larger than 0.05. The rest of the five significant age groups are 25-34, 35-45, 55-64,65-74, 75, and over. There are three age groups(55-64, 65-74,75, and over) that have negative coefficients, which are -0.1671,-0.5929, and -1.4789, respectively. There are two age groups, 25-34 and 35-45, with a positive coefficient(0.056498 and 0.032735). When people are in above two age groups, people's odds will increase by 0.056498 and 0.032735, respectively.

```{r,include=FALSE,echo=TRUE}
# create a bar plot with different education levels and (smoking, non-smoking)
ggplot(ss2, aes(x = education_level, fill = smoking_or_not)) +
    geom_bar(position = "fill") +
    theme_classic() + 
  labs(x="education level",
       title= "Figure 4: The distribution of individuals smoking given six education levels  ",
       subtitle = "Changes in individuals smoking given different education levels ",
       cex.names = 0.9)+
  theme_bw()+theme(axis.text.x = element_text(angle = 20, hjust = 1, size=10,color="darkred"))
```

Figure 4 shows that the number of people in education levels such as not stated, valid skip and trade school, college choose to smoke is approximately the same, and most among all six education levels. The baseline for education levels is high school. However, when an individual is a refusal to answer the education level, the odds of people will decrease by -9.4556. 

```{r,include=FALSE,echo=TRUE}
# create a bar plot with number of children in one household and (smoking, non-smoking)
ggplot(ss2, aes(x = number_of_children, fill = smoking_or_not)) +
    geom_bar(position = "fill") +
    theme_classic() + 
  labs(x="number of children in one household ",
       title= "Figure 5: The distribution of individuals smoking given number of children",
       subtitle = "Changes in individuals smoking given different numbers of children ",
       cex.names = 0.9)+
  theme_bw()+theme(axis.text.x = element_text(angle = 30, hjust = 1, size=10,color="darkred"))
```

Figure 5 shows that a number of households with 0,1,2,4 and more kids choose to smoke have approximately the same number of households with three kids choosing to smoke has a smaller percentage. The baseline for the number of children is four or more children. In the logistic regression model result, when one household has three kids, the odds of smoking are decreased by -0.153934.   

Overall, people who have 0-2 children or four and more children are more likely to smoke. People who are extremely stressed, not at all stressful, not stated, quite a bit of stressful or a bit stressful are more likely to smoke. People whose education_level is a trade school, college, university, or valid skip are more likely to smoke. Furthermore, people who have 15-45 years of age are more likely to smoke. 


##Discussion

From the logistic regression model result, we could tell that young people and students are more likely to smoke. They have different stress levels. Students are stressed by a huge amount of homework, scores, and their parents. Then they are easy to smoke for relaxation. Young adults who have massive works to do or stress raising their children are also more likely to smoke. 

Our model can be used to provide information on the smoking prevention program. If we know the necessary information of a person, then we can use the logistic regression model to predict the odds of individuals who smoke. When an individual has more odds of smoking, the smoking prevention program can send an email that includes information about stopping smoking to these people. People are less likely to control themselves from smoking. If we can stay away from smoking initially, we will have less risk of having the disease. 

As we mentioned before, the dataset is from the 2016 GSS. Therefore, we can not guarantee a precise prediction according to the past data. We will collect some more recent data and future data to do a new logistic regression model in future work, we can predict with a more precise model result. Ten provinces within Canada stratify the data. There are limitations to geographic focus. We can only predict Canadians smoke or not using the 2016 GSS data. However, we would like to let people focus on the health problem should not only in Canada but also worldwide. As a result, we will also collect more detailed survey data from worldwide to predict people will smoke or not in different countries.
Furthermore, we would like to select more predictor variables. For example, we may choose more predictor variables such as income, occupation, and parents in the household. The more predictor variables we use in the logistic regression model, the more precise prediction we will get. 








##Reference
-GSS 2016 Dataset: https://sda-artsci-utoronto-ca.myaccess.library.utoronto.ca/cgi-bin/sda/hsda2?setupfile=harcsda4&datasetname=gss30&ui=2&action=subset
-H. Wickham. ggplot2: Elegant Graphics for Data Analysis. Springer-Verlag New York, 2016.
-R Core Team (2020). R: A language and environment for statistical computing. R Foundation for Statistical Computing, Vienna, Austria. URL
  https://www.R-project.org/.
-Wickham et al., (2019). Welcome to the tidyverse. Journal of Open Source Software, 4(43), 1686, https://doi.org/10.21105/joss.01686
-T. Lumley (2020) "survey: analysis of complex survey samples". R package version 4.0.
-T. Lumley (2004) Analysis of complex survey samples. Journal of Statistical Software 9(1): 1-19
-T. Lumley (2010) Complex Surveys: A Guide to Analysis Using R. John Wiley and Sons.
